---
title: "Analisis Satisfaccion de Cliente"
author: "Juan Carlos Ibanez"
date: "2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = '../working/Customer Satisfaction')
```

## 1. Introduction

Customer Experience and Satisfaction is a key point in today business.

For this analysis we have a survey of customer satifaction questions for a set of customers in 5 different companies.

The dataset is a sample from a real-life bigger one. 

As the responsable for customer experience in one of this companies, you are asked to solve several business questions as:   

  1. What are they key questions to ask to a customer in order to understand his opinion
  2. How many questions do we need to ask in order to register a good view of the customer experience with our customer
  3. What are the "dimensions" of the customer satisfaction
  4. Which dimensions are the most important ones
  5. Are those dimensions different for young customers (quantile 50% of age), than for olders

## 2. Data Cleansing and Preparation


```{r}
#
#
#
```


## 3. Exploratory Data Analysis

```{r}
#
#
#
```

## 4. What are the most important areas/dimensions in customer satisfaction in the mobile telecomunications business?

```{r}
#
#
#
```

### 4.1. Principal Component Analysis

Try a Principal Component Analysis to study the dimensions of customer satisfaction
```{r}
#
#
#
```

How many PCs will you select to keep a 70% of the variability?
```{r}
#
#
#
```


### 4.2. Factor Analysis

How many factors do we need in order to keep a 70% of the original information
```{r}
#
#
#
```

Explain the factors you have found
```{r}
#
#
#
```
How many dimensions for the customer satisfaccion will you choose
```{r}
#
#
#
```

## 5. What is the impact of each dimension in recomendation?

Calculate the regression between the factors and the vatiable RECOMENDACION  

### 5.1. Multiple Regression Analysis

```{r}
#
#
#
```
#### 5.2. What dimensions are more important for global satisfaction?

```{r}
#
#
#
```
