---
title: "Descriptive Multivariate Analysis: countries by socio-economic-health"
author: "Máster en Data Science y Big Data en Finanzas"
date: 'AFI, 2022'
output:
  html_document: 
    css: my-theme.css
    theme: cerulean
    highlight: tango
    number_sections: no
    toc: no
    toc_depth: 1
  pdf_document:
    css: my-theme.css
    theme: cerulean
    highlight: tango
    number_sections: yes
    toc: yes
    toc_depth: 1
editor_options:
  chunk_output_type: console
---


```{r global_options, include=T, echo = F}
knitr::opts_chunk$set(echo = T, warning=FALSE, message=FALSE)
```

```{r, echo=FALSE}
htmltools::img(src = knitr::image_uri(file.path("AfiLogo.jpg")), 
               alt = 'logo', 
               style = 'position:absolute; top:0; right:0; padding:10px;',
               width="173",
               height="77")
```

# Motivation

Can we organize the world's countries by variables we are interested in?

For instance, the World Bank organizes countries in 4 income groups using just the Gross National Income (GNI) per capita

But what if we are interested in health? And in much more variables?

In this case study, we will analyze countries using socio-economic and health indicators that determine the overall development of the world

```{r}
library(tidyverse)
library(countrycode)
library(rworldmap)
library(GGally)
```

## The dataset 

The dataset is obtained through the notebook "WHO_preprocessing.Rmd", which creates a file named "WorldData2022.txt". There you can find explanation for the following World Health Organization (WHO) variables:

  - Life expectancy at birth (years)
  - Infant mortality rate (probability of dying between birth and age 1 per 1000 live births)
  - Population (in thousands) total
  - Ambient air pollution (Annual PM2.5 [ug/m3])
  - Estimates of rates of homicides per 100 000 population
  - Hospital beds (per 10 000 population)
  - Total expenditure on health as a percentage of gross domestic product
  - Percentage of individuals using the Internet
  - Physicians density (per 1000 population)
  
and also from the following The World Happiness Report (WHR) variables:

  - Happiness
  - GDP per capita
  - Healthy life expectancy
  - Social support
  - Perceived freedom
  - Generosity
  - Perception of corruption

```{r}
WHO = read.table("WorldData2022.txt", sep=",")  
```

## Descriptive analysis

  
```{r}
dim(WHO)
str(WHO)
head(WHO)
tail(WHO)
summary(WHO)
```

Out of the 200 countries in the world, we have almost complete information (20 variables) for 151 countries: 

  - Diverse variables (socio-economic and health)

  - Just a few missing values
  
One country is duplicated:

```{r}
WHO$Country[duplicated(WHO$Country)]
WHO = WHO[-33,]
```

The output of the tools depend directly on the selected variables for the input. This selection is subjective and depends on the application's objective

```{r}
X = WHO %>% dplyr::select(-Continent,-Region,-Country, -COUNTRY,-region,-country,-Population) 
```
  
Take care: we are going to omit rows with NAs

```{r}
dim(WHO)[1]
dim(na.omit(X))[1]

id.complete = complete.cases(X)
names = WHO$Country[id.complete]
continent = WHO$Continent[id.complete]
population = WHO$Population[id.complete]

X = X[id.complete,]
row.names(X)=names
```
  
# Basic Data Analysis

Tables

```{r}
table(WHO$Continent)
```

Better with a barplot:

```{r}
WHO %>% ggplot(aes(x=reorder(Continent, Continent, length))) +geom_bar(aes(fill=Continent)) + 
  labs(caption="Countries per continent",
       x = "", y = "")+ theme(legend.position="none")
```
  
Basic maps:

```{r}
# select your favourite variable here:
map = WHO %>% dplyr::select(Country, LifeExpectancy)

#Convert the country code into iso3c using the function countrycode()
map$country = countrycode(map$Country, 'country.name', 'iso3c')

#Create data object supporting the map
matched <- joinCountryData2Map(map, joinCode = "ISO3",
                               nameJoinColumn = "country")
#Draw the map
mapCountryData(matched,nameColumnToPlot="LifeExpectancy",missingCountryCol = "white",
               borderCol = "#C7D9FF",
               catMethod = "pretty", colourPalette = "topo",
               mapTitle = c("Life Expectancy by Country"), lwd=1)
```

Which is the country with the lowest young population?

And the country with the highest senior population? Are they the same?
  
# Basic descriptive measures 

Enter here the vector of means

```{r}
xbar = colMeans(X)  
xbar
```

Enter here the covariance matrix

```{r}
S = cov(X)   
S
```

Enter here the correlation matrix

```{r}
R = cor(X)  
R
```

heterogeneous data units:

```{r}
det(S)
sum(diag(S))
```

highly correlated data:

```{r}
det(R)
sum(diag(R))
```

## Relations in dimension 1

Univariate analysis for all p=14 variables

```{r}
boxplot(X, col="blue", las=2)
```

Scale or not to scale?

```{r}
boxplot(scale(X), las=2, col="darkblue")
```

## Relations in dimension 2

Bivariate analysis (scatter plots), in total 78

Multiple scatter plot: all relations, 78, in dimension 2

```{r}
ggcorr(X, label = T)
```

Insights?

For instance, happiness is highly correlated with GDP, social support, and population using internet. And negatively correlated with infant mortality rate

## Relations in dimension 3

In total 286

In general, no way to obtain information...

But sometimes we can...

For instance, this is 4D using a scatterplot (2D)

```{r}
ggplot(WHO, aes(x=HealthExpenditure, y=LifeExpectancy, group=Continent, size=Population, color=happiness)) + 
  geom_point(alpha=0.9) + geom_smooth(method=lm,se=F, size=0.3) +
  facet_wrap(~ Continent) +
  scale_color_gradient(low="red", high="green") +
  theme_minimal()+ theme(legend.position="none") + 
  labs(title = "World countries: life expectancy vs health expenditure", subtitle="(color denotes happiness, size denotes population)",caption="Source: World Health Organization",
       x = "Expenditure on health (percentage of GDP)", y = "Life expectancy at birth (in years)")

```

Is wealth a determinant of happiness?

```{r}
WHO %>% ggplot(aes(x=gdp, y=happiness, group=Continent, size=Population, color=region)) + 
  geom_point(alpha=0.75) + geom_smooth(method=lm,se=F, size=0.3) +
  facet_wrap(~ Continent) +
  theme_minimal()+ theme(legend.position="none") + 
  labs(title = "Happiness vs wealth by continent", subtitle="(size denotes population, color denotes region)",caption="Source: World Happiness Report",
       x = "GDP (log)", y = "Happiness Score")
```


# Multivariate standardization

This is very useful to compute distances considering correlations: Mahalanobis distance

```{r}
S_x <- cov(X)
iS <- solve(S_x)
e <- eigen(iS)
V <- e$vectors
B <- V %*% diag(sqrt(e$values)) %*% t(V)
Xtil <- scale(X,scale = FALSE)
X.S <- Xtil %*% B
```

Mahalanobis distances between each country and the world's center:

```{r}
muhat = colMeans(X)
dM2 <- mahalanobis(X,muhat,S)  # Mahalanobis distances (square)
dM2

plot(sort(dM2,index.return=TRUE)$x,pch=19,col=grey.colors(n=length(dM2),start=0,end=.75),xlab="",ylab="",main="Mahalanobis distances (square)")

sort(dM2)
```

Closest countries to Spain:

```{r}
dM2 <- mahalanobis(X,t(X["Spain",]),S)
head(sort(dM2))
```

# Motivation to PCA

What is the happiest country?

What is the country with the highest life expectancy? Is it the same?

What is the country with the lowest IMR?

Can we rank the countries considering all the variables at the same time?

## From dimension X to dimension 2

```{r}
pca = prcomp(X, scale=T)
summary(pca)
```

With 2 components we explain 63% of variability

```{r}
library(factoextra)

fviz_screeplot(pca, addlabels = TRUE)
```

First component?

```{r}
barplot(pca$rotation[,1], las=2, col="darkblue")
```

It is a measure of global development (positive loads on good indicators and negative ones on the bad variables)

The most important indicators are: Internet, gdp, LifeExpectancy, IMR, happiness, etc.

Note the sum of the squared loadings (eigenvectors) is equal to 1

```{r}
sum(pca$rotation[,1]^2)
```

That means squared loadings are easier to interpret than the loadings

I.e. they are like percentages (numbers between 0 and 1), but we loose the sign

Let's plot squared loadings instead. 

They are called contribution of variables to components

```{r}
fviz_contrib(pca, choice = "var", axes = 1)
```

The red dashed line on the graph above indicates the expected average contribution 

If the contribution of the variables were uniform, the expected value would be 1/length(variables) = 1/14 = 7%

## Rank the countries by the first component (scores)

```{r}
names[order(pca$x[,1])][1:10]
names[order(pca$x[,1], decreasing=T)][1:10]
```

Second component?

```{r}
barplot(pca$rotation[,2], las=2, col="darkblue")
```

Contribution of variables to second component

```{r}
fviz_contrib(pca, choice = "var", axes = 2)
```

# Plot the first two scores

```{r}
data.frame(z1=pca$x[,1],z2=pca$x[,2]) %>% 
  ggplot(aes(z1,z2,label=names,color=continent)) + geom_point(size=0) +
  labs(title="First two principal components (scores)", x="PC1", y="PC2") + #guides(color=guide_legend(title="HDI"))+
  theme_bw() +theme(legend.position="bottom") + geom_text(size=3, hjust=0.6, vjust=0, check_overlap = TRUE) 
# The two first PCs seem independent

```

Conclusions?

Map:

```{r}
# Map our PCA index in a map:
map = data.frame(country=names, value=pca$x[,1])
#Convert the country code into iso3c using the function countrycode()
map$country = countrycode(map$country, 'country.name', 'iso3c')
#Create data object supporting the map
matched <- joinCountryData2Map(map, joinCode = "ISO3",
                               nameJoinColumn = "country")
#Draw the map
mapCountryData(matched,nameColumnToPlot="value",missingCountryCol = "white",
               addLegend = FALSE, borderCol = "#C7D9FF",
               catMethod = "pretty", colourPalette = "terrain",
               mapTitle = c("PCA1 by Country"), lwd=1)

# zoom in Africa
mapCountryData(matched, nameColumnToPlot="value", ,missingCountryCol = "white",
               addLegend = FALSE, borderCol = "#C7D9FF",
               mapRegion = "Africa", mapTitle = c("PCA1 by Country"), lwd=1, 
               catMethod = "quantile", colourPalette = "terrain")



```

