---
title: "Notebook to download socio-economic data in the world"
author: "Máster en Data Science y Big Data en Finanzas"
date: 'AFI, 2022'
output:
  html_document: 
    css: my-theme.css
    theme: cerulean
    highlight: tango
    number_sections: no
    toc: no
    toc_depth: 1
  pdf_document:
    css: my-theme.css
    theme: cerulean
    highlight: tango
    number_sections: yes
    toc: yes
    toc_depth: 1
editor_options:
  chunk_output_type: console
---


```{r global_options, include=T, echo = F}
knitr::opts_chunk$set(echo = T, warning=FALSE, message=FALSE)
```

```{r, echo=FALSE}
htmltools::img(src = knitr::image_uri(file.path("AfiLogo.jpg")), 
               alt = 'logo', 
               style = 'position:absolute; top:0; right:0; padding:10px;',
               width="173",
               height="77")
```

Load and prepare data from the World Health Organization (WHO) connecting through its API

Here we can browse WHO data by topic:
  https://apps.who.int/gho/data/node.main
  

```{r}
# devtools::install_github("pierucci/rgho@devel")
library(rgho)  #Api WHO
library(tidyverse)
library(countrycode)
library(rworldmap)
library(GGally)
```

Let's start downloading variables from WHO
  
```{r}
# Create a data frame with all countries in the world
WHO = dplyr::arrange(
  rgho:::to_data_frame(
    get_gho_codes(dimension = "COUNTRY")
  ),
  Label)

# Regions:
#  AFR  Africa
#  AMR  Americas
#  EMR  Eastern Mediterranean
#  EUR  Europe
#  SEAR South-East Asia
#  WPR  Western Pacific

WHO = WHO %>% rename(Country=Label, COUNTRY=ID)

# WHOSIS_000001: Life expectancy at birth (years)
var = get_gho_data("WHOSIS_000001") %>% mutate(LifeExpectancy=Numeric) %>%
  group_by(COUNTRY) %>% filter(YEAR==max(YEAR,na.rm=T)) %>% arrange(SEX) %>% slice(1) %>% 
  dplyr::select(COUNTRY, LifeExpectancy)
WHO = merge(WHO, var, by="COUNTRY", all=T)
dim(WHO)

# MDG_0000000001: Infant mortality rate (probability of dying between birth and age 1 per 1000 live births)
var = get_gho_data("MDG_0000000001") %>% mutate(IMR = Numeric) %>% 
  group_by(COUNTRY) %>% filter(YEAR==max(YEAR,na.rm=T)) %>% arrange(SEX) %>% slice(1) %>% 
  dplyr::select(COUNTRY, IMR) 
WHO = merge(WHO, var, by="COUNTRY", all=T)

# # WHS9_86: Population (in thousands) total
# var = get_gho_data("WHS9_86") %>% mutate(Population = Numeric) %>%
#   group_by(COUNTRY) %>% filter(YEAR==max(YEAR,na.rm=T)) %>% slice(1) %>%
#   dplyr::select(COUNTRY, Population)
# WHO = merge(WHO, var, by="COUNTRY", all=T)

# Population is now taken from World Bank data bank, available through tidyr
var = filter(world_bank_pop, indicator=="SP.POP.TOTL")[,c(1,20)]
names(var) = c("COUNTRY", "Population")
var$Population = var$Population/1000
WHO = merge(WHO, var, by="COUNTRY", all=T)

# SDGPM25 : Annual mean levels of fine particulate matter (PM2.5)
var = get_gho_data("SDGPM25") %>% mutate(AirPollution = Numeric) %>%
  group_by(COUNTRY) %>% filter(YEAR==max(YEAR,na.rm=T)) %>% slice(1) %>%
  dplyr::select(COUNTRY, AirPollution)
WHO = merge(WHO, var, by="COUNTRY", all=T)

# VIOLENCE_HOMICIDERATE: Estimates of rates of homicides per 100 000 population
var = get_gho_data("VIOLENCE_HOMICIDERATE") %>% mutate(HomicidesRate = Numeric) %>% 
  group_by(COUNTRY) %>% filter(YEAR==max(YEAR,na.rm=T)) %>% slice(1) %>% 
  dplyr::select(COUNTRY, HomicidesRate) 
WHO = merge(WHO, var, by="COUNTRY", all=T)

# WHS6_102: Hospital beds (per 10 000 population)
var = get_gho_data("WHS6_102") %>% mutate(HospitalBeds = Numeric) %>% 
  group_by(COUNTRY) %>% filter(YEAR==max(YEAR,na.rm=T)) %>% slice(1) %>% 
  dplyr::select(COUNTRY, HospitalBeds) 
WHO = merge(WHO, var, by="COUNTRY", all=T)

# WHS7_143: Total expenditure on health as a percentage of gross domestic product
var = get_gho_data("WHS7_143") %>% mutate(HealthExpenditure = Numeric) %>% 
  group_by(COUNTRY) %>% filter(YEAR==max(YEAR,na.rm=T)) %>% slice(1) %>% 
  dplyr::select(COUNTRY, HealthExpenditure) 
WHO = merge(WHO, var, by="COUNTRY", all=T)

# ITU_ICT_1: Percentage of individuals using the Internet
var = get_gho_data("ITU_ICT_1") %>% mutate(Internet = Numeric) %>% 
  group_by(COUNTRY) %>% filter(YEAR==max(YEAR,na.rm=T)) %>% slice(1) %>% 
  dplyr::select(COUNTRY, Internet) 
WHO = merge(WHO, var, by="COUNTRY", all=T)

# HRH_26: Physicians density (per 1000 population)
var = get_gho_data("HRH_26") %>% mutate(Physicians = Numeric) %>% 
  group_by(COUNTRY) %>% filter(YEAR==max(YEAR,na.rm=T)) %>% slice(1) %>% 
  dplyr::select(COUNTRY, Physicians) 
WHO = merge(WHO, var, by="COUNTRY", all=T)


# Names for Continents and Regions, and remove countries with no Region
WHO$Continent = countrycode(WHO$Country, "country.name", destination="continent")
WHO$Region = countrycode(WHO$Country, "country.name", destination="region")
WHO = WHO %>% drop_na(Continent,Region)

WHO = WHO %>% 
  dplyr::select(Country,COUNTRY,Continent, Region, everything())
```

Now, we are going to merge data from The World Happiness Report (WHR) https://worldhappiness.report/ed/2020/

This report ranks cities and countries based on a survey about life evaluations. 

The WHR considers 6 underlying determinants for happiness: GDP per capita, healthy life expectancy, social support, perceived freedom, generosity, and perception of corruption.

More information here: https://happiness-report.s3.amazonaws.com/2020/WHR20_Ch2_Statistical_Appendix.pdf

```{r}
happiness <- read.csv('WHR20_Data.csv', stringsAsFactors=T)
names(happiness)[c(1,2,3,7,8,9,10,11,12)] = c("country", "region", "happiness", "gdp", "social_support", "life_expectancy", "perceived_freedom", "generosity", "perception_of_corruption")
happiness = happiness[,c(1,2,3,7,8,10,11,12)]

happiness$COUNTRY = countrycode(happiness$country, 'country.name', 'iso3c')

df = merge(WHO, happiness, by="COUNTRY")
```

Write the file to load it later

```{r}
write.table(df, "WorldData2022.txt", sep=",")
```
